---
title: "SARS-CoV-2 dataset: Differential expression analysis"
author: "Alberto Valdeolivas: alberto.valdeolivas@bioquant.uni-heidelberg.de; Date: "
date: "14/04/2020"
output: github_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## Introduction

The present script takes the RNAseq data from the study *"SARS-CoV-2 launches* 
*a unique transcriptional signature from in vitro, ex vivo, and in vivo systems"* 

<https://www.biorxiv.org/content/10.1101/2020.03.24.004655v1>

<https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147507>

It performs a differential expression analysis to compare:

+ Human lung epithelial cells (NHBE): mock treated vs infected with SARS-CoV-2. 

+ A549 alveolar cancer cell line: mock treated vs infected with SARS-CoV-2.   

+ A549 cell line does not express ACE2, the receptor used by SARS-CoV-2 to 
penetrate into human cells. Therefore A549 were also transduced with ACE2 and 
then mock treated or infected with SARS-CoV-2

+ Calu-3 human lung epithelial cancer cell line: mock treated vs 
infected with SARS-CoV-2.   

We used the *DESeq2* R package.

## Getting Started

We first load the required libraries. 

```{r, message=FALSE}
library(dplyr)
library(DESeq2)
```

We also read the raw counts from the original experiment (GSE147507)

```{r}
## Raw counts table
GSE147507_raw_counts <- 
    read.csv("RawData/GSE147507_RawReadCounts_Human.tsv", sep = "\t")
```


## NHBE mock treated vs infected with SARS-CoV-2 

We first select the Series 1, which corresponds to independent biological 
triplicates of primary human lung epithelium (NHBE) that were either mock 
treated or infected with SARS-CoV-2. 

```{r, message=FALSE}
## We select series 1 as described above.
count_NHBEvsCOV2_df <- GSE147507_raw_counts[,c(2:7)]
row.names(count_NHBEvsCOV2_df) <- GSE147507_raw_counts$X
```

We create a dataframe describing the experimental design. 

```{r}
targets_NHBEvsCOV2 <- 
    as.data.frame(matrix(NA,length(names(count_NHBEvsCOV2_df)),1))
names(targets_NHBEvsCOV2) <- c("condition")
row.names(targets_NHBEvsCOV2) <- names(count_NHBEvsCOV2_df)
targets_NHBEvsCOV2$condition <- 
    gsub("Series1_", "", row.names(targets_NHBEvsCOV2))
targets_NHBEvsCOV2$condition <- 
    factor(gsub("_[1-3]$", "", targets_NHBEvsCOV2$condition))
targets_NHBEvsCOV2
```

Then, we perform the differential expression analysis with *DESeq2*

```{r, message=FALSE}
## Create deseq2 object
dds_NHBEvsCOV2 <- 
    DESeqDataSetFromMatrix(countData = as.matrix(count_NHBEvsCOV2_df), 
    colData = targets_NHBEvsCOV2, design = ~ condition)

## Set control
dds_NHBEvsCOV2$condition <- relevel(dds_NHBEvsCOV2$condition, 
    ref = levels(targets_NHBEvsCOV2$condition)[1])

## Carry out diff exp
dds_NHBEvsCOV2 <- DESeq(dds_NHBEvsCOV2)
```

We finally save the table with the results of the analysis and the normalised
counts

```{r}
## See the comparisons carried out
comparison_NHBEvsCOV2 <- resultsNames(dds_NHBEvsCOV2)

## Get results table
results_NHBEvsCOV2 <- 
    results(dds_NHBEvsCOV2, name=comparison_NHBEvsCOV2[2])

## Save the object
saveRDS(results_NHBEvsCOV2, file="IntermediateFiles/dds_results_NHBEvsCOV2.rds")

## Extract normalised counts data
dds_NHBEvsCOV2 <- estimateSizeFactors(dds_NHBEvsCOV2)
norm_counts_NHBEvsCOV2 <- counts(dds_NHBEvsCOV2, normalized = TRUE)
saveRDS(norm_counts_NHBEvsCOV2, 
    file="IntermediateFiles/counts_norm_NHBEvsCOV2.rds")
```

## A549 mock treated vs infected with SARS-CoV-2 

Then, we select the Series 2, which corresponds to independent biological 
triplicates of the A549 alveolar cancer cell line that were either mock 
treated or infected with SARS-CoV-2. 

```{r, message=FALSE}
## We select series 2 as described above.
count_A549vsCOV2_df <- GSE147507_raw_counts[,c(8:13)]
row.names(count_A549vsCOV2_df) <- GSE147507_raw_counts$X
```

We create a dataframe describing the experimental design. 

```{r}
targets_A549vsCOV2 <- 
    as.data.frame(matrix(NA,length(names(count_A549vsCOV2_df)),1))
names(targets_A549vsCOV2) <- c("condition")
row.names(targets_A549vsCOV2) <- names(count_A549vsCOV2_df)
targets_A549vsCOV2$condition <- 
    gsub("Series2_", "", row.names(targets_A549vsCOV2))
targets_A549vsCOV2$condition <- 
    factor(gsub("_[1-3]$", "", targets_A549vsCOV2$condition))
targets_A549vsCOV2
```

Then, we perform the differential expression analysis with *DESeq2*

```{r, message=FALSE}
## Create deseq2 object
dds_A549vsCOV2 <- 
    DESeqDataSetFromMatrix(countData = as.matrix(count_A549vsCOV2_df), 
    colData = targets_A549vsCOV2, design = ~ condition)

## Set control
dds_A549vsCOV2$condition <- relevel(dds_A549vsCOV2$condition, 
    ref = levels(targets_A549vsCOV2$condition)[1])

## Carry out diff exp
dds_A549vsCOV2 <- DESeq(dds_A549vsCOV2)
```

We finally save the table with the results of the analysis and the normalised 
counts:

```{r}
## See the comparisons carried out
comparison_A549vsCOV2 <- resultsNames(dds_A549vsCOV2)

## Get results table
results_A549vsCOV2 <- 
    results(dds_A549vsCOV2, name=comparison_A549vsCOV2[2])

## Save the object
saveRDS(results_A549vsCOV2, file="IntermediateFiles/dds_results_A549vsCOV2.rds")

## Extract normalised counts data
dds_A549vsCOV2 <- estimateSizeFactors(dds_A549vsCOV2)
norm_counts_A549vsCOV2 <- counts(dds_A549vsCOV2, normalized = TRUE)
saveRDS(norm_counts_A549vsCOV2, 
    file="IntermediateFiles/counts_norm_A549vsCOV2.rds")
```

## A549 transduced with ACE2 mock treated vs infected with SARS-CoV-2 

A549 does not express the ACE2 receptor used by SARS-CoV-2 to infect cells. 
Therefore, A549 were also transduced with a vector expressing human ACE2, 
were also mock treated or infected with SARS-CoV-2.

```{r, message=FALSE}
## We select series 6 as described above.
count_A549ACE2vsCOV2_df <- GSE147507_raw_counts[,c(28:33)]
row.names(count_A549ACE2vsCOV2_df) <- GSE147507_raw_counts$X
```

We create a dataframe describing the experimental design. 

```{r}
targets_A549ACE2vsCOV2 <- 
    as.data.frame(matrix(NA,length(names(count_A549ACE2vsCOV2_df)),1))
names(targets_A549ACE2vsCOV2) <- c("condition")
row.names(targets_A549ACE2vsCOV2) <- names(count_A549ACE2vsCOV2_df)
targets_A549ACE2vsCOV2$condition <- 
    gsub("Series2_", "", row.names(targets_A549ACE2vsCOV2))
targets_A549ACE2vsCOV2$condition <- 
    factor(gsub("_[1-3]$", "", targets_A549ACE2vsCOV2$condition))
targets_A549ACE2vsCOV2
```

Then, we perform the differential expression analysis with *DESeq2*

```{r, message=FALSE}
## Create deseq2 object
dds_A549ACE2vsCOV2 <- 
    DESeqDataSetFromMatrix(countData = as.matrix(count_A549ACE2vsCOV2_df), 
    colData = targets_A549ACE2vsCOV2, design = ~ condition)

## Set control
dds_A549ACE2vsCOV2$condition <- relevel(dds_A549ACE2vsCOV2$condition, 
    ref = levels(targets_A549ACE2vsCOV2$condition)[1])

## Carry out diff exp
dds_A549ACE2vsCOV2 <- DESeq(dds_A549ACE2vsCOV2)
```

We finally save the table with the results of the analysis and the normalised 
counts:

```{r}
## See the comparisons carried out
comparison_A549ACE2vsCOV2 <- resultsNames(dds_A549ACE2vsCOV2)

## Get results table
results_A549ACE2vsCOV2 <- 
    results(dds_A549ACE2vsCOV2, name=comparison_A549ACE2vsCOV2[2])

## Save the object
saveRDS(results_A549ACE2vsCOV2, file="IntermediateFiles/dds_results_A549ACE2vsCOV2.rds")

## Extract normalised counts data
dds_A549ACE2vsCOV2 <- estimateSizeFactors(dds_A549ACE2vsCOV2)
norm_counts_A549ACE2vsCOV2 <- counts(dds_A549ACE2vsCOV2, normalized = TRUE)
saveRDS(norm_counts_A549ACE2vsCOV2, 
    file="IntermediateFiles/counts_norm_A549ACE2vsCOV2.rds")
```


## CALU-3 mock treated vs infected with SARS-CoV-2 

We then select the Series 7, which corresponds to independent biological 
triplicates of a human lung epithelium cancer cell line named Calu-3, that were 
either mock treated or infected with SARS-CoV-2. 

```{r, message=FALSE}
## We select series 7 as described above.
count_CALU3vsCOV2_df <- GSE147507_raw_counts[,c(34:39)]
row.names(count_CALU3vsCOV2_df) <- GSE147507_raw_counts$X
```

We create a dataframe describing the experimental design. 

```{r}
targets_CALU3vsCOV2 <- 
    as.data.frame(matrix(NA,length(names(count_CALU3vsCOV2_df)),1))
names(targets_CALU3vsCOV2) <- c("condition")
row.names(targets_CALU3vsCOV2) <- names(count_CALU3vsCOV2_df)
targets_CALU3vsCOV2$condition <- 
    gsub("Series1_", "", row.names(targets_CALU3vsCOV2))
targets_CALU3vsCOV2$condition <- 
    factor(gsub("_[1-3]$", "", targets_CALU3vsCOV2$condition))
targets_CALU3vsCOV2
```

Then, we perform the differential expression analysis with *DESeq2*

```{r, message=FALSE}
## Create deseq2 object
dds_CALU3vsCOV2 <- 
    DESeqDataSetFromMatrix(countData = as.matrix(count_CALU3vsCOV2_df), 
    colData = targets_CALU3vsCOV2, design = ~ condition)

## Set control
dds_CALU3vsCOV2$condition <- relevel(dds_CALU3vsCOV2$condition, 
    ref = levels(targets_CALU3vsCOV2$condition)[1])

## Carry out diff exp
dds_CALU3vsCOV2 <- DESeq(dds_CALU3vsCOV2)
```

We finally save the table with the results of the analysis and the normalised
counts

```{r}
## See the comparisons carried out
comparison_CALU3vsCOV2 <- resultsNames(dds_CALU3vsCOV2)

## Get results table
results_CALU3vsCOV2 <- 
    results(dds_CALU3vsCOV2, name=comparison_CALU3vsCOV2[2])

## Save the object
saveRDS(results_CALU3vsCOV2, file="IntermediateFiles/dds_results_CALU3vsCOV2.rds")

## Extract normalised counts data
dds_CALU3vsCOV2 <- estimateSizeFactors(dds_CALU3vsCOV2)
norm_counts_CALU3vsCOV2 <- counts(dds_CALU3vsCOV2, normalized = TRUE)
saveRDS(norm_counts_CALU3vsCOV2, 
    file="IntermediateFiles/counts_norm_CALU3vsCOV2.rds")
```

## Session Info Details

```{r, echo=FALSE, eval=TRUE}
sessionInfo()
```

