---
title: "SARS-CoV-2 dataset: Running Progeny and Dorothea"
author: "Alberto Valdeolivas: alberto.valdeolivas@bioquant.uni-heidelberg.de; Date: "
date: "14/04/2020"
output: github_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## Introduction

The present script deal with the RNAseq data from the study *"SARS-CoV-2 launches* 
*a unique transcriptional signature from in vitro, ex vivo, and in vivo systems"* 

<https://www.biorxiv.org/content/10.1101/2020.03.24.004655v1>

<https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147507>

It uses the differential analysis results generated in the previous script to
run **Progeny** and **Dorothea** for the four cell lines under study:

+ Human lung epithelial cells (NHBE): mock treated vs infected with SARS-CoV-2. 

+ A549 alveolar cancer cell line: mock treated vs infected with SARS-CoV-2.   

+ A549 cell line does not express ACE2, the receptor used by SARS-CoV-2 to 
penetrate into human cells. Therefore A549 were also transduced with ACE2 and 
then mock treated or infected with SARS-CoV-2

+ Calu-3 human lung epithelial cancer cell line: mock treated vs 
infected with SARS-CoV-2.   

## Getting Started

We first load the required libraries. 

```{r, message=FALSE}
library(devtools)
# install_github("saezlab/progeny")
# install_github("saezlab/dorothea")
library(progeny)
library(dorothea)
library(tibble)
library(tidyr)
library(dplyr)
library(ggplot2)
library(pheatmap)
```

We also read the differential expression results and normalised counts from the 
previous script.

```{r}
## Differential expression table
dds_NHBEvsCOV2 <- readRDS("IntermediateFiles/dds_results_NHBEvsCOV2.rds")
dds_A549vsCOV2 <- readRDS("IntermediateFiles/dds_results_A549vsCOV2.rds")
dds_A549ACE2vsCOV2 <- 
  readRDS("IntermediateFiles/dds_results_A549ACE2vsCOV2.rds") 
dds_CALU3vsCOV2 <- readRDS("IntermediateFiles/dds_results_CALU3vsCOV2.rds")

norm_counts_NHBEvsCOV2 <- 
    readRDS("IntermediateFiles/counts_norm_NHBEvsCOV2.rds")
norm_counts_A549vsCOV2 <- 
    readRDS("IntermediateFiles/counts_norm_A549vsCOV2.rds")
norm_counts_A549ACE2vsCOV2 <- 
    readRDS("IntermediateFiles/counts_norm_A549ACE2vsCOV2.rds")
norm_counts_CALU3vsCOV2 <- 
    readRDS("IntermediateFiles/counts_norm_CALU3vsCOV2.rds")
```

## Pathway activity with Progeny

We first estimate the pathway activity using the **Progeny** package. 

### NHBE mock treated vs infected with SARS-CoV-2 

We first compute **Progeny** scores per every sample (with the replicates) using 
the normalised counts. We display the results in a Heatmap.

```{r}
pathways_NHBEvsCOV2_counts <- progeny(norm_counts_NHBEvsCOV2, scale=TRUE, 
    organism="Human", top = 100)
activities_NHBEvsCOV2 <- as.vector(pathways_NHBEvsCOV2_counts)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

progenyBreaks <- c(seq(min(activities_NHBEvsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(activities_NHBEvsCOV2)/paletteLength, 
    max(activities_NHBEvsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
progeny_hmap <- pheatmap(t(pathways_NHBEvsCOV2_counts),fontsize=14, 
    fontsize_row = 10, fontsize_col = 10, 
    color=myColor, breaks = progenyBreaks, 
    main = "PROGENy (100)", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

We can see that the different replicates cluster together and clear different
activity patterns between the infected and the mock treated NHBE lines. To 
evaluate the significance and sign of these pathway activities, we can run again
**Progeny** using the statistic from the differential express analysis.

```{r, message=FALSE}
dds_NHBEvsCOV2_df <- as.data.frame(dds_NHBEvsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") 

pathways_NHBEvsCOV2_zscore <- t(progeny(as.matrix(dds_NHBEvsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = TRUE))
colnames(pathways_NHBEvsCOV2_zscore) <- "NES"

## I also need to run progeny in such a way to have values between 1 and -1 to
## use as CARNIVAL input
pathways_NHBEvsCOV2_zscore_inputCarnival <- 
  t(progeny(as.matrix(dds_NHBEvsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = FALSE))
colnames(pathways_NHBEvsCOV2_zscore_inputCarnival) <- "Activity"
```

We now display the normalized enrichment scores (NES) in a bar plot. 

```{r, message=FALSE}
pathways_NHBEvsCOV2_zscore_df <- as.data.frame(pathways_NHBEvsCOV2_zscore) %>% 
    rownames_to_column(var = "Pathway") %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(Pathway = factor(Pathway))
```
```{r, dpi = 300}
ggplot(pathways_NHBEvsCOV2_zscore_df,aes(x = reorder(Pathway, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways")
```

### A549 mock treated vs infected with SARS-CoV-2 

We reproduce the progeny activity analysis for A549 cell line.

We first compute **Progeny** scores per every sample (with the replicates) using 
the normalised counts. We display the results in a Heatmap.

```{r}
pathways_A549vsCOV2_counts <- progeny(norm_counts_A549vsCOV2, scale=TRUE, 
    organism="Human", top = 100)
activities_A549vsCOV2 <- as.vector(pathways_A549vsCOV2_counts)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

progenyBreaks <- c(seq(min(activities_A549vsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(activities_A549vsCOV2)/paletteLength, 
    max(activities_A549vsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
progeny_hmap <- pheatmap(t(pathways_A549vsCOV2_counts),fontsize=14, 
    fontsize_row = 10, fontsize_col = 10, 
    color=myColor, breaks = progenyBreaks, 
    main = "PROGENy (100)", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

We can see that the different replicates cluster together and clear different
activity patterns between the infected and the mock treated A549 lines. To 
evaluate the significance and sign of these pathway activities, we can run again
**Progeny** using the statistic from the differential express analysis.

```{r, message=FALSE}
dds_A549vsCOV2_df <- as.data.frame(dds_A549vsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") 

pathways_A549vsCOV2_zscore <- t(progeny(as.matrix(dds_A549vsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = TRUE))
colnames(pathways_A549vsCOV2_zscore) <- "NES"

## I also need to run progeny in such a way to have values between 1 and -1 to
## use as CARNIVAL input
pathways_A549vsCOV2_zscore_inputCarnival <- 
  t(progeny(as.matrix(dds_A549vsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = FALSE))
colnames(pathways_A549vsCOV2_zscore_inputCarnival) <- "Activity"
```

We now display the normalized enrichment scores (NES) in a bar plot. 

```{r, message=FALSE}
pathways_A549vsCOV2_zscore_df <- as.data.frame(pathways_A549vsCOV2_zscore) %>% 
    rownames_to_column(var = "Pathway") %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(Pathway = factor(Pathway))
```
```{r, message=FALSE, dpi=300}
ggplot(pathways_A549vsCOV2_zscore_df,aes(x = reorder(Pathway, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways")
```

### A549 transduced with ACE2 mock treated vs infected with SARS-CoV-2 

We reproduce the progeny activity analysis for A549 transduced with ACE2 cell 
line.

We first compute **Progeny** scores per every sample (with the replicates) using 
the normalised counts. We display the results in a Heatmap.

```{r}
pathways_A549ACE2vsCOV2_counts <- progeny(norm_counts_A549ACE2vsCOV2, 
    scale=TRUE, organism="Human", top = 100)
activities_A549ACE2vsCOV2 <- as.vector(pathways_A549ACE2vsCOV2_counts)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

progenyBreaks <- c(seq(min(activities_A549ACE2vsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(activities_A549ACE2vsCOV2)/paletteLength, 
    max(activities_A549ACE2vsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
progeny_hmap <- pheatmap(t(pathways_A549ACE2vsCOV2_counts),fontsize=14, 
    fontsize_row = 10, fontsize_col = 10, 
    color=myColor, breaks = progenyBreaks, 
    main = "PROGENy (100)", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

We can see that the different replicates cluster together and clear different
activity patterns between the infected and the mock treated A549 lines 
(transduced with ACE2). To evaluate the significance and sign of these pathway 
activities, we can run again **Progeny** using the statistic from the 
differential express analysis.

```{r, message=FALSE}
dds_A549ACE2vsCOV2_df <- as.data.frame(dds_A549ACE2vsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") 

pathways_A549ACE2vsCOV2_zscore <- t(progeny(as.matrix(dds_A549ACE2vsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = TRUE))
colnames(pathways_A549ACE2vsCOV2_zscore) <- "NES"

## I also need to run progeny in such a way to have values between 1 and -1 to
## use as CARNIVAL input
pathways_A549ACE2vsCOV2_zscore_inputCarnival <- 
  t(progeny(as.matrix(dds_A549ACE2vsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = FALSE))
colnames(pathways_A549ACE2vsCOV2_zscore_inputCarnival) <- "Activity"
```

We now display the normalized enrichment scores (NES) in a bar plot. 

```{r, message=FALSE}
pathways_A549ACE2vsCOV2_zscore_df <- as.data.frame(pathways_A549ACE2vsCOV2_zscore) %>% 
    rownames_to_column(var = "Pathway") %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(Pathway = factor(Pathway))
```
```{r, message=FALSE, dpi=300}
ggplot(pathways_A549ACE2vsCOV2_zscore_df,aes(x = reorder(Pathway, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways")
```

### CALU-3 mock treated vs infected with SARS-CoV-2 

We first compute **Progeny** scores per every sample (with the replicates) using 
the normalised counts. We display the results in a Heatmap.

```{r}
pathways_CALU3vsCOV2_counts <- progeny(norm_counts_CALU3vsCOV2, scale=TRUE, 
    organism="Human", top = 100)
activities_CALU3vsCOV2 <- as.vector(pathways_CALU3vsCOV2_counts)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

progenyBreaks <- c(seq(min(activities_CALU3vsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(activities_CALU3vsCOV2)/paletteLength, 
    max(activities_CALU3vsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
progeny_hmap <- pheatmap(t(pathways_CALU3vsCOV2_counts),fontsize=14, 
    fontsize_row = 10, fontsize_col = 10, 
    color=myColor, breaks = progenyBreaks, 
    main = "PROGENy (100)", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

We can see that the different replicates cluster together and clear different
activity patterns between the infected and the mock treated CALU-3 lines. To 
evaluate the significance and sign of these pathway activities, we can run again
**Progeny** using the statistic from the differential express analysis.

```{r, message=FALSE}
dds_CALU3vsCOV2_df <- as.data.frame(dds_CALU3vsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") 

pathways_CALU3vsCOV2_zscore <- t(progeny(as.matrix(dds_CALU3vsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = TRUE))
colnames(pathways_CALU3vsCOV2_zscore) <- "NES"

## I also need to run progeny in such a way to have values between 1 and -1 to
## use as CARNIVAL input
pathways_CALU3vsCOV2_zscore_inputCarnival <- 
  t(progeny(as.matrix(dds_CALU3vsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = FALSE))
colnames(pathways_CALU3vsCOV2_zscore_inputCarnival) <- "Activity"
```

We now display the normalized enrichment scores (NES) in a bar plot. 

```{r, message=FALSE}
pathways_CALU3vsCOV2_zscore_df <- as.data.frame(pathways_CALU3vsCOV2_zscore) %>% 
    rownames_to_column(var = "Pathway") %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(Pathway = factor(Pathway))
```
```{r, dpi = 300}
ggplot(pathways_CALU3vsCOV2_zscore_df,aes(x = reorder(Pathway, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways")
```

### Comparison: Pathway activity for SARS-CoV infection in different lines

We finally plot together the pathway activities for the 4 different 
cell lines. 

```{r}
AllInfections_Pathways <- bind_rows(pathways_NHBEvsCOV2_zscore_df,
    pathways_A549vsCOV2_zscore_df, pathways_A549ACE2vsCOV2_zscore_df,
    pathways_CALU3vsCOV2_zscore_df) %>% 
    tibble::add_column(
        CellLine = c(rep("NHBE",14),rep("A549",14), 
                  rep("A549_ACE2",14), rep("CALU-3",14)))
```
```{r, dpi=300}
ggplot(AllInfections_Pathways,aes(x = Pathway, y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    facet_wrap(~CellLine, ncol = 2, nrow = 2) +
    scale_fill_gradient2(low = "darkblue", high = "darkred", 
        mid = "white", midpoint = 0) + 
    theme_classic() +
    theme(axis.title = element_text(face = "bold", size = 14),
        axis.text.x = 
            element_text(angle = 90, hjust = 1, size =8),
        axis.text.y = element_text(size =8),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways") 
```

## Transcription Factor activity with Dorothea and Viper

Now, we estimate the transcription factor (TF) activity using the **dorothea** 
package. We select Dorothea interactions with confidence level A,B and C. 

```{r}
## We load Dorothea Regulons
data(dorothea_hs, package = "dorothea")
regulons <- dorothea_hs %>%
  dplyr::filter(confidence %in% c("A", "B","C"))
```

It is to note that for **Dorothea**, we proceed the other way around than for 
**Progeny**. We have many TFs so, we cannot visualize all of them in the same 
heatmap. That is why, we first compute the TF activity using the statistcs from 
the differential expression analysis. This will allows us to select the TFs 
whose activity varies the most between the mock treated and infected samples. 

### NHBE mock treated vs infected with SARS-CoV-2 

We first run Viper using the statistic from the different expression analysis.

```{r}
dds_NHBEvsCOV2_stat <-  as.data.frame(dds_NHBEvsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_NHBEvsCOV2_stat <- 
    dorothea::run_viper(as.matrix(dds_NHBEvsCOV2_stat), regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, nes = TRUE))
```

We now display the top 25 normalized enrichment scores (NES) for the TF in a 
bar plot. 

```{r}
tf_activities_NHBEvsCOV2_top25 <- tf_activities_NHBEvsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(25, wt = abs(NES)) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))
``` 
```{r, dpi=300}
ggplot(tf_activities_NHBEvsCOV2_top25,aes(x = reorder(GeneID, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Transcription Factors")
```

We now compute TFs activities per every sample (with the replicates) using 
the normalised counts. We display the results of the previous 25 TFs in a 
Heatmap.

```{r}
tf_activities_NHBEvsCOV2_counts <- 
    dorothea::run_viper(norm_counts_NHBEvsCOV2, regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, method = c("scale")))

tf_activities_NHBEvsCOV2_counts_filter <- tf_activities_NHBEvsCOV2_counts %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::filter(GeneID %in% tf_activities_NHBEvsCOV2_top25$GeneID) %>%
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_NHBEvsCOV2 <- as.vector(tf_activities_NHBEvsCOV2_counts_filter)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

dorotheaBreaks <- c(seq(min(tf_activities_NHBEvsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(tf_activities_NHBEvsCOV2)/paletteLength, 
    max(tf_activities_NHBEvsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
dorothea_hmap <- pheatmap(tf_activities_NHBEvsCOV2_counts_filter,
    fontsize=14, fontsize_row = 8, fontsize_col = 8, 
    color=myColor, breaks = dorotheaBreaks,
    main = "Dorothea ABC", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

### A549 mock treated vs infected with SARS-CoV-2 

We first run Viper using the statistic from the different expression analysis.

```{r}
dds_A549vsCOV2_stat <-  as.data.frame(dds_A549vsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549vsCOV2_stat <- 
    dorothea::run_viper(as.matrix(dds_A549vsCOV2_stat), regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, nes = TRUE))
```

We now display the top 25 normalized enrichment scores (NES) for the TF in a 
bar plot. 

```{r}
tf_activities_A549vsCOV2_top25 <- tf_activities_A549vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(25, wt = abs(NES)) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))
```   
```{r, dpi=300}
ggplot(tf_activities_A549vsCOV2_top25,aes(x = reorder(GeneID, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Transcription Factors")
```

We now compute TFs activities per every sample (with the replicates) using 
the normalised counts. We display the results of the previous 25 TFs in a 
Heatmap.

```{r}
tf_activities_A549vsCOV2_counts <- 
    dorothea::run_viper(norm_counts_A549vsCOV2, regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, method = c("scale")))

tf_activities_A549vsCOV2_counts_filter <- tf_activities_A549vsCOV2_counts %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::filter(GeneID %in% tf_activities_A549vsCOV2_top25$GeneID) %>%
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549vsCOV2 <- as.vector(tf_activities_A549vsCOV2_counts_filter)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

dorotheaBreaks <- c(seq(min(tf_activities_A549vsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(tf_activities_A549vsCOV2)/paletteLength, 
    max(tf_activities_A549vsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
dorothea_hmap <- pheatmap(tf_activities_A549vsCOV2_counts_filter,
    fontsize=14, fontsize_row = 8, fontsize_col = 8, 
    color=myColor, breaks = dorotheaBreaks,
    main = "Dorothea ABC", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

### A549 transduced with ACE2 mock treated vs infected with SARS-CoV-2 

We first run Viper using the statistic from the different expression analysis.

```{r}
dds_A549ACE2vsCOV2_stat <-  as.data.frame(dds_A549ACE2vsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549ACE2vsCOV2_stat <- 
    dorothea::run_viper(as.matrix(dds_A549ACE2vsCOV2_stat), regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, nes = TRUE))
```

We now display the top 25 normalized enrichment scores (NES) for the TF in a 
bar plot. 

```{r}
tf_activities_A549ACE2vsCOV2_top25 <- tf_activities_A549ACE2vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(25, wt = abs(NES)) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))
```   
```{r, dpi=300}
ggplot(tf_activities_A549ACE2vsCOV2_top25,aes(x = reorder(GeneID, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Transcription Factors")
```

We now compute TFs activities per every sample (with the replicates) using 
the normalised counts. We display the results of the previous 25 TFs in a 
Heatmap.

```{r}
tf_activities_A549ACE2vsCOV2_counts <- 
    dorothea::run_viper(norm_counts_A549ACE2vsCOV2, regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, method = c("scale")))

tf_activities_A549ACE2vsCOV2_counts_filter <- tf_activities_A549ACE2vsCOV2_counts %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::filter(GeneID %in% tf_activities_A549ACE2vsCOV2_top25$GeneID) %>%
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549ACE2vsCOV2 <- as.vector(tf_activities_A549ACE2vsCOV2_counts_filter)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

dorotheaBreaks <- c(seq(min(tf_activities_A549ACE2vsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(tf_activities_A549ACE2vsCOV2)/paletteLength, 
    max(tf_activities_A549ACE2vsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
dorothea_hmap <- pheatmap(tf_activities_A549ACE2vsCOV2_counts_filter,
    fontsize=14, fontsize_row = 8, fontsize_col = 8, 
    color=myColor, breaks = dorotheaBreaks,
    main = "Dorothea ABC", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

### CALU-3 mock treated vs infected with SARS-CoV-2 

We first run Viper using the statistic from the different expression analysis.

```{r}
dds_CALU3vsCOV2_stat <-  as.data.frame(dds_CALU3vsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_CALU3vsCOV2_stat <- 
    dorothea::run_viper(as.matrix(dds_CALU3vsCOV2_stat), regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, nes = TRUE))
```

We now display the top 25 normalized enrichment scores (NES) for the TF in a 
bar plot. 

```{r}
tf_activities_CALU3vsCOV2_top25 <- tf_activities_CALU3vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(25, wt = abs(NES)) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))
``` 
```{r, dpi=300}
ggplot(tf_activities_CALU3vsCOV2_top25,aes(x = reorder(GeneID, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Transcription Factors")
```

We now compute TFs activities per every sample (with the replicates) using 
the normalised counts. We display the results of the previous 25 TFs in a 
Heatmap.

```{r}
tf_activities_CALU3vsCOV2_counts <- 
    dorothea::run_viper(norm_counts_CALU3vsCOV2, regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, method = c("scale")))

tf_activities_CALU3vsCOV2_counts_filter <- tf_activities_CALU3vsCOV2_counts %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::filter(GeneID %in% tf_activities_CALU3vsCOV2_top25$GeneID) %>%
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_CALU3vsCOV2 <- as.vector(tf_activities_CALU3vsCOV2_counts_filter)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

dorotheaBreaks <- c(seq(min(tf_activities_CALU3vsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(tf_activities_CALU3vsCOV2)/paletteLength, 
    max(tf_activities_CALU3vsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
dorothea_hmap <- pheatmap(tf_activities_CALU3vsCOV2_counts_filter,
    fontsize=14, fontsize_row = 8, fontsize_col = 8, 
    color=myColor, breaks = dorotheaBreaks,
    main = "Dorothea ABC", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```


### Comparison: TF activity for SARS-CoV infection in different lines

We finally plot together the pathway activities for the 4 different 
cell lines.

```{r, message=FALSE, warning=FALSE}
AllInfections_TFs <- bind_rows(tf_activities_NHBEvsCOV2_top25,
    tf_activities_A549vsCOV2_top25, tf_activities_A549ACE2vsCOV2_top25,
    tf_activities_CALU3vsCOV2_top25) %>% 
    tibble::add_column(
        CellLine = c(rep("NHBE",25),rep("A549",25),
            rep("A549_ACE2",25),rep("CALU-3",25)))
```
```{r, dpi=300}
ggplot(AllInfections_TFs,aes(x = GeneID, y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    facet_wrap(~CellLine,drop= FALSE, nrow = 2, ncol=2) +
    scale_fill_gradient2(low = "darkblue", high = "darkred", 
        mid = "white", midpoint = 0) + 
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 90, hjust = 1, size =5),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) 
```

The previous plot is not very clear. Then and since we aim at focusing on the 
effect SARS-CoV-2 infection regardless of the cell line, we selected those TFs 
that are among the top active/unactive in every cell line.

```{r}
### I am going to select the top 75 up & down separated

## NHBE
tf_activities_NHBEvsCOV2_top75_up <- tf_activities_NHBEvsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(75, wt = NES) %>%
    dplyr::arrange(desc(NES)) %>% 
    pull(GeneID)

tf_activities_NHBEvsCOV2_top75_down <- tf_activities_NHBEvsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(75, wt = -NES) %>%
    dplyr::arrange(NES) %>% 
     pull(GeneID)

## A549
tf_activities_A549vsCOV2_top75_up <- tf_activities_A549vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(75, wt = NES) %>%
    dplyr::arrange(desc(NES)) %>% 
     pull(GeneID)

tf_activities_A549vsCOV2_top75_down <- tf_activities_A549vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(75, wt = -NES) %>%
    dplyr::arrange(NES) %>% 
    pull(GeneID)

## A549 transduced with ACE2
tf_activities_A549ACE2vsCOV2_top75_up <- tf_activities_A549ACE2vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(75, wt = NES) %>%
    dplyr::arrange(desc(NES)) %>% 
    pull(GeneID)

tf_activities_A549ACE2vsCOV2_top75_down <- tf_activities_A549ACE2vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(75, wt = -NES) %>%
    dplyr::arrange(NES) %>% 
    pull(GeneID)

## CALU3 
tf_activities_CALU3vsCOV2_top75_up <- tf_activities_CALU3vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(75, wt = NES) %>%
    dplyr::arrange(desc(NES)) %>% 
    pull(GeneID)

tf_activities_CALU3vsCOV2_top75_down <- tf_activities_CALU3vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(75, wt = -NES) %>%
    dplyr::arrange(NES) %>% 
    pull(GeneID)

TFscommon_up <- 
  intersect(
    intersect(
      intersect(
        tf_activities_NHBEvsCOV2_top75_up, 
        tf_activities_A549vsCOV2_top75_up),
        tf_activities_A549ACE2vsCOV2_top75_up), 
        tf_activities_CALU3vsCOV2_top75_up)

TFscommon_down <- intersect(
  intersect(
    intersect(
      tf_activities_NHBEvsCOV2_top75_down, 
      tf_activities_A549vsCOV2_top75_down),
      tf_activities_A549ACE2vsCOV2_top75_down), 
      tf_activities_CALU3vsCOV2_top75_down)

TFscommon_up_down <- c(TFscommon_up,TFscommon_down)

TFscommon_up_down_NES <- bind_rows(tf_activities_NHBEvsCOV2_stat %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "GeneID") %>%
  dplyr::filter(GeneID %in% TFscommon_up_down) %>% 
  tibble::add_column(CellLine = c(rep("NHBE",14))), 
  tf_activities_A549vsCOV2_stat %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "GeneID") %>%
  dplyr::filter(GeneID %in% TFscommon_up_down) %>% 
  tibble::add_column(CellLine = c(rep("A549",14))),
  tf_activities_A549ACE2vsCOV2_stat %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "GeneID") %>%
  dplyr::filter(GeneID %in% TFscommon_up_down) %>% 
  tibble::add_column(CellLine = c(rep("A549ACE2",14))),
  tf_activities_CALU3vsCOV2_stat %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "GeneID") %>%
  dplyr::filter(GeneID %in% TFscommon_up_down) %>% 
  tibble::add_column(CellLine = c(rep("CALU3",14)))
  ) %>% 
    dplyr::rename(NES = "stat") 
```
```{r, dpi=300}
ggplot(TFscommon_up_down_NES,aes(x = GeneID, y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    facet_wrap(~CellLine,drop= FALSE, nrow = 2, ncol=2) +
    scale_fill_gradient2(low = "darkblue", high = "darkred", 
        mid = "white", midpoint = 0) + 
    theme_classic() +
    theme(axis.title = element_text(face = "bold", size = 14),
        axis.text.x = 
            element_text(angle = 90, hjust = 1, size =8),
        axis.text.y = element_text(size =8),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
        xlab("Transcription Factors")
```

### Saving Results 

```{r, eval=FALSE}
saveRDS(pathways_NHBEvsCOV2_counts, 
    file = "IntermediateFiles/pathways_NHBEvsCOV2_counts.rds")
saveRDS(pathways_NHBEvsCOV2_zscore, 
    file = "IntermediateFiles/pathways_NHBEvsCOV2_zscore.rds")
saveRDS(pathways_NHBEvsCOV2_zscore_inputCarnival,
    file = "IntermediateFiles/pathways_NHBEvsCOV2_zscore_inputCarnival.rds")
saveRDS(tf_activities_NHBEvsCOV2_stat, 
    file = "IntermediateFiles/tf_activities_NHBEvsCOV2_stat.rds")
saveRDS(tf_activities_NHBEvsCOV2_counts,
    file = "IntermediateFiles/tf_activities_NHBEvsCOV2_counts.rds")

saveRDS(pathways_A549vsCOV2_counts, 
    file = "IntermediateFiles/pathways_A549vsCOV2_counts.rds")
saveRDS(pathways_A549vsCOV2_zscore,
    file = "IntermediateFiles/pathways_A549vsCOV2_zscore.rds")
saveRDS(pathways_A549vsCOV2_zscore_inputCarnival,
    file = "IntermediateFiles/pathways_A549vsCOV2_zscore_inputCarnival.rds")
saveRDS(tf_activities_A549vsCOV2_stat, 
    file = "IntermediateFiles/tf_activities_A549vsCOV2_stat.rds")
saveRDS(tf_activities_A549vsCOV2_counts, 
    file = "IntermediateFiles/tf_activities_A549vsCOV2_counts.rds")

saveRDS(pathways_A549ACE2vsCOV2_counts, 
    file = "IntermediateFiles/pathways_A549ACE2vsCOV2_counts.rds")
saveRDS(pathways_A549ACE2vsCOV2_zscore,
    file = "IntermediateFiles/pathways_A549ACE2vsCOV2_zscore.rds")
saveRDS(pathways_A549ACE2vsCOV2_zscore_inputCarnival,
    file = "IntermediateFiles/pathways_A549ACE2vsCOV2_zscore_inputCarnival.rds")
saveRDS(tf_activities_A549ACE2vsCOV2_stat, 
    file = "IntermediateFiles/tf_activities_A549ACE2vsCOV2_stat.rds")
saveRDS(tf_activities_A549ACE2vsCOV2_counts, 
    file = "IntermediateFiles/tf_activities_A549ACE2vsCOV2_counts.rds")

saveRDS(pathways_CALU3vsCOV2_counts, 
    file = "IntermediateFiles/pathways_CALU3vsCOV2_counts.rds")
saveRDS(pathways_CALU3vsCOV2_zscore, 
    file = "IntermediateFiles/pathways_CALU3vsCOV2_zscore.rds")
saveRDS(pathways_CALU3vsCOV2_zscore_inputCarnival,
    file = "IntermediateFiles/pathways_CALU3vsCOV2_zscore_inputCarnival.rds")
saveRDS(tf_activities_CALU3vsCOV2_stat, 
    file = "IntermediateFiles/tf_activities_CALU3vsCOV2_stat.rds")
saveRDS(tf_activities_CALU3vsCOV2_counts,
    file = "IntermediateFiles/tf_activities_CALU3vsCOV2_counts.rds")
```


## Session Info Details

```{r, echo=FALSE, eval=TRUE}
sessionInfo()
```

