---
title: "SARS-CoV-2 vs RSV vs HPIV3: Differential expression analysis"
author: "Alberto Valdeolivas: alberto.valdeolivas@bioquant.uni-heidelberg.de; Date: "
date: "30/04/2020"
output: github_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## Introduction

The present script takes the RNAseq data from the study *"SARS-CoV-2 launches* 
*a unique transcriptional signature from in vitro, ex vivo, and in vivo systems"* 

<https://www.biorxiv.org/content/10.1101/2020.03.24.004655v1>

<https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147507>

The goal of this set of scripts is to compare the transcriptional effect of 
different viral infections: SARS-CoV-2, Respiratory syncytial virus (RSV) and 
Human parainfluenza virus type 3 (HPIV3). To do so, we first perform a 
differential expression analysis to compare the following samples:

+ A549 alveolar cancer cell line: mock treated vs infected with SARS-CoV-2.   

+ A549 alveolar cancer cell line: mock treated vs infected with RSV. 

+ A549 alveolar cancer cell line: mock treated vs infected with HPIV3.

The differential expression analysis was performed using the *DESeq2* R package.

## Getting Started

We first load the required libraries. 

```{r, message=FALSE}
library(dplyr)
library(DESeq2)
```

We also read the raw counts from the original experiment (GSE147507)

```{r}
## Raw counts table
GSE147507_raw_counts <- 
    read.csv("../RawData/GSE147507_RawReadCounts_Human.tsv", sep = "\t")
```


## A549 mock treated vs infected with SARS-CoV-2 

We first select the Series 5, which corresponds to independent biological 
triplicates of alveolar cancer cell line (A549) that were either mock 
treated or infected with SARS-CoV-2. 

```{r, message=FALSE}
## We select series 5 as described above.
count_A549vsCOV2_df <- GSE147507_raw_counts[,c(22:27)]
row.names(count_A549vsCOV2_df) <- GSE147507_raw_counts$X
```

We create a dataframe describing the experimental design. 

```{r}
targets_A549vsCOV2 <- 
    as.data.frame(matrix(NA,length(names(count_A549vsCOV2_df)),1))
names(targets_A549vsCOV2) <- c("condition")
row.names(targets_A549vsCOV2) <- names(count_A549vsCOV2_df)
targets_A549vsCOV2$condition <- 
    gsub("Series5_", "", row.names(targets_A549vsCOV2))
targets_A549vsCOV2$condition <- 
    factor(gsub("_[1-3]$", "", targets_A549vsCOV2$condition))
targets_A549vsCOV2
```

Then, we perform the differential expression analysis with *DESeq2*

```{r, message=FALSE}
## Create deseq2 object
dds_A549vsCOV2 <- 
    DESeqDataSetFromMatrix(countData = as.matrix(count_A549vsCOV2_df), 
    colData = targets_A549vsCOV2, design = ~ condition)

## Set control
dds_A549vsCOV2$condition <- relevel(dds_A549vsCOV2$condition, 
    ref = levels(targets_A549vsCOV2$condition)[1])

## Carry out diff exp
dds_A549vsCOV2 <- DESeq(dds_A549vsCOV2)
```

We finally save the table with the results of the analysis and the normalised
counts

```{r}
## See the comparisons carried out
comparison_A549vsCOV2 <- resultsNames(dds_A549vsCOV2)

## Get results table
results_A549vsCOV2 <- 
    results(dds_A549vsCOV2, name=comparison_A549vsCOV2[2])

## Save the object
saveRDS(results_A549vsCOV2,
    file="IntermediateFiles/dds_results_A549vsCOV2.rds")

## Extract normalised counts data
dds_A549vsCOV2 <- estimateSizeFactors(dds_A549vsCOV2)
norm_counts_A549vsCOV2 <- counts(dds_A549vsCOV2, normalized = TRUE)
saveRDS(norm_counts_A549vsCOV2, 
    file="IntermediateFiles/counts_norm_A549vsCOV2.rds")
```


## A549 mock treated vs infected with RSV.

We now select the Series 8, which corresponds to independent biological 
triplicates of alveolar cancer cell line (A549) that were either mock 
treated, infected with RSV or infected with HPIV3. We focus here in pairwise 
comparison between mock treated and infected with RSV

```{r, message=FALSE}
## We select series 5 as described above.
count_A549vsRSV_df <- GSE147507_raw_counts[,c(40:45)]
row.names(count_A549vsRSV_df) <- GSE147507_raw_counts$X
```

We create a dataframe describing the experimental design. 

```{r}
targets_A549vsRSV <- 
    as.data.frame(matrix(NA,length(names(count_A549vsRSV_df)),1))
names(targets_A549vsRSV) <- c("condition")
row.names(targets_A549vsRSV) <- names(count_A549vsRSV_df)
targets_A549vsRSV$condition <- 
    gsub("Series8_", "", row.names(targets_A549vsRSV))
targets_A549vsRSV$condition <- 
    factor(gsub("_[1-3]$", "", targets_A549vsRSV$condition))
targets_A549vsRSV
```

Then, we perform the differential expression analysis with *DESeq2*

```{r, message=FALSE}
## Create deseq2 object
dds_A549vsRSV <- 
    DESeqDataSetFromMatrix(countData = as.matrix(count_A549vsRSV_df), 
    colData = targets_A549vsRSV, design = ~ condition)

## Set control
dds_A549vsRSV$condition <- relevel(dds_A549vsRSV$condition, 
    ref = levels(targets_A549vsRSV$condition)[1])

## Carry out diff exp
dds_A549vsRSV <- DESeq(dds_A549vsRSV)
```

We finally save the table with the results of the analysis and the normalised
counts

```{r}
## See the comparisons carried out
comparison_A549vsRSV <- resultsNames(dds_A549vsRSV)

## Get results table
results_A549vsRSV <- 
    results(dds_A549vsRSV, name=comparison_A549vsRSV[2])

## Save the object
saveRDS(results_A549vsRSV,
    file="IntermediateFiles/dds_results_A549vsRSV.rds")

## Extract normalised counts data
dds_A549vsRSV <- estimateSizeFactors(dds_A549vsRSV)
norm_counts_A549vsRSV <- counts(dds_A549vsRSV, normalized = TRUE)
saveRDS(norm_counts_A549vsRSV, 
    file="IntermediateFiles/counts_norm_A549vsRSV.rds")
```

## A549 mock treated vs infected with HPIV3.

We again select the Series 8, which corresponds to independent biological 
triplicates of alveolar cancer cell line (A549) that were either mock 
treated, infected with RSV or infected with HPIV3. We focus now in pairwise 
comparison between mock treated and infected with HPIV3.

```{r, message=FALSE}
## We select series 5 as described above.
count_A549vsHPIV3_df <- GSE147507_raw_counts[,c(40:42,46:48)]
row.names(count_A549vsHPIV3_df) <- GSE147507_raw_counts$X
```

We create a dataframe describing the experimental design. 

```{r}
targets_A549vsHPIV3 <- 
    as.data.frame(matrix(NA,length(names(count_A549vsHPIV3_df)),1))
names(targets_A549vsHPIV3) <- c("condition")
row.names(targets_A549vsHPIV3) <- names(count_A549vsHPIV3_df)
targets_A549vsHPIV3$condition <- 
    gsub("Series8_", "", row.names(targets_A549vsHPIV3))
targets_A549vsHPIV3$condition <- 
    factor(gsub("_[1-3]$", "", targets_A549vsHPIV3$condition))
targets_A549vsHPIV3$condition <- 
    factor(gsub("_HPIV3", "HPIV3", targets_A549vsHPIV3$condition))
targets_A549vsHPIV3
```

Then, we perform the differential expression analysis with *DESeq2*

```{r, message=FALSE}
## Create deseq2 object
dds_A549vsHPIV3 <- 
    DESeqDataSetFromMatrix(countData = as.matrix(count_A549vsHPIV3_df), 
    colData = targets_A549vsHPIV3, design = ~ condition)

## Set control
dds_A549vsHPIV3$condition <- relevel(dds_A549vsHPIV3$condition, 
    ref = levels(targets_A549vsHPIV3$condition)[1])

## Carry out diff exp
dds_A549vsHPIV3 <- DESeq(dds_A549vsHPIV3)
```

We finally save the table with the results of the analysis and the normalised
counts

```{r}
## See the comparisons carried out
comparison_A549vsHPIV3 <- resultsNames(dds_A549vsHPIV3)

## Get results table
results_A549vsHPIV3 <- 
    results(dds_A549vsHPIV3, name=comparison_A549vsHPIV3[2])

## Save the object
saveRDS(results_A549vsHPIV3,
    file="IntermediateFiles/dds_results_A549vsHPIV3.rds")

## Extract normalised counts data
dds_A549vsHPIV3 <- estimateSizeFactors(dds_A549vsHPIV3)
norm_counts_A549vsHPIV3 <- counts(dds_A549vsHPIV3, normalized = TRUE)
saveRDS(norm_counts_A549vsHPIV3, 
    file="IntermediateFiles/counts_norm_A549vsHPIV3.rds")
```


## Session Info Details

```{r, echo=FALSE, eval=TRUE}
sessionInfo()
```

