---
title: "SARS-CoV-2 vs RSV vs HPIV3: running Progeny and Dorothea"
author: "Alberto Valdeolivas: alberto.valdeolivas@bioquant.uni-heidelberg.de; Date: "
date: "30/04/2020"
output: github_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## Introduction

The present script takes the RNAseq data from the study *"SARS-CoV-2 launches* 
*a unique transcriptional signature from in vitro, ex vivo, and in vivo systems"* 

<https://www.biorxiv.org/content/10.1101/2020.03.24.004655v1>

<https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147507>

The goal of this set of scripts is to compare the transcriptional effect of 
different viral infections: SARS-CoV-2, Respiratory syncytial virus (RSV) and 
Human parainfluenza virus type 3 (HPIV3). In this script we use the differential 
analysis results generated in the previous script torun **Progeny** and 
**Dorothea** for the conditions under study:

+ A549 alveolar cancer cell line: mock treated vs infected with SARS-CoV-2.   

+ A549 alveolar cancer cell line: mock treated vs infected with RSV. 

+ A549 alveolar cancer cell line: mock treated vs infected with HPIV3.

## Getting Started

We first load the required libraries. 

```{r, message=FALSE}
library(devtools)
# install_github("saezlab/progeny")
# install_github("saezlab/dorothea")
library(progeny)
library(dorothea)
library(tibble)
library(tidyr)
library(dplyr)
library(ggplot2)
library(pheatmap)
```

We also read the differential expression results and normalised counts from the 
previous script.

```{r}
## Differential expression table
dds_A549vsCOV2 <- readRDS("IntermediateFiles/dds_results_A549vsCOV2.rds")
dds_A549vsRSV <- readRDS("IntermediateFiles/dds_results_A549vsRSV.rds")
dds_A549vsHPIV3 <- readRDS("IntermediateFiles/dds_results_A549vsHPIV3.rds")

## Normalised counts
norm_counts_A549vsCOV2 <- 
    readRDS("IntermediateFiles/counts_norm_A549vsCOV2.rds")
norm_counts_A549vsRSV <- 
    readRDS("IntermediateFiles/counts_norm_A549vsRSV.rds")
norm_counts_A549vsHPIV3 <- 
    readRDS("IntermediateFiles/counts_norm_A549vsHPIV3.rds")
```

## Pathway activity with Progeny

We first estimate the pathway activity using the **Progeny** package. 

### A549 mock treated vs infected with SARS-CoV-2 

We perform the progeny activity analysis for A549 cell line infected with 
SARS-CoV-2.

We first compute **Progeny** scores per every sample (with the replicates) using 
the normalised counts. We display the results in a Heatmap.

```{r}
pathways_A549vsCOV2_counts <- progeny(norm_counts_A549vsCOV2, scale=TRUE, 
    organism="Human", top = 100)
activities_A549vsCOV2 <- as.vector(pathways_A549vsCOV2_counts)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

progenyBreaks <- c(seq(min(activities_A549vsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(activities_A549vsCOV2)/paletteLength, 
    max(activities_A549vsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
progeny_hmap <- pheatmap(t(pathways_A549vsCOV2_counts),fontsize=14, 
    fontsize_row = 10, fontsize_col = 10, 
    color=myColor, breaks = progenyBreaks, 
    main = "PROGENy (100)", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

We can see that the different replicates cluster together and clear different
activity patterns between the infected and the mock treated A549 lines. To 
evaluate the significance and sign of these pathway activities, we can run again
**Progeny** using the statistic from the differential express analysis.

```{r, message=FALSE}
dds_A549vsCOV2_df <- as.data.frame(dds_A549vsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") 

pathways_A549vsCOV2_zscore <- t(progeny(as.matrix(dds_A549vsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = TRUE))
colnames(pathways_A549vsCOV2_zscore) <- "NES"

## I also need to run progeny in such a way to have values between 1 and -1 to
## use as CARNIVAL input
pathways_A549vsCOV2_zscore_inputCarnival <- 
  t(progeny(as.matrix(dds_A549vsCOV2_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = FALSE))
colnames(pathways_A549vsCOV2_zscore_inputCarnival) <- "Activity"
```

We now display the normalized enrichment scores (NES) in a bar plot. 

```{r, message=FALSE}
pathways_A549vsCOV2_zscore_df <- as.data.frame(pathways_A549vsCOV2_zscore) %>% 
    rownames_to_column(var = "Pathway") %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(Pathway = factor(Pathway))
```
```{r, message=FALSE, dpi=300}
ggplot(pathways_A549vsCOV2_zscore_df,aes(x = reorder(Pathway, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways")
```

### A549 mock treated vs infected with RSV

We perform the progeny activity analysis for A549 cell line infected with RSV.

We first compute **Progeny** scores per every sample (with the replicates) using 
the normalised counts. We display the results in a Heatmap.

```{r}
pathways_A549vsRSV_counts <- progeny(norm_counts_A549vsRSV, scale=TRUE, 
    organism="Human", top = 100)
activities_A549vsRSV <- as.vector(pathways_A549vsRSV_counts)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

progenyBreaks <- c(seq(min(activities_A549vsRSV), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(activities_A549vsRSV)/paletteLength, 
    max(activities_A549vsRSV), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
progeny_hmap <- pheatmap(t(pathways_A549vsRSV_counts),fontsize=14, 
    fontsize_row = 10, fontsize_col = 10, 
    color=myColor, breaks = progenyBreaks, 
    main = "PROGENy (100)", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

We can see that the different replicates cluster together and clear different
activity patterns between the infected and the mock treated A549 lines. To 
evaluate the significance and sign of these pathway activities, we can run again
**Progeny** using the statistic from the differential express analysis.

```{r, message=FALSE}
dds_A549vsRSV_df <- as.data.frame(dds_A549vsRSV) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") 

pathways_A549vsRSV_zscore <- t(progeny(as.matrix(dds_A549vsRSV_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = TRUE))
colnames(pathways_A549vsRSV_zscore) <- "NES"

## I also need to run progeny in such a way to have values between 1 and -1 to
## use as CARNIVAL input
pathways_A549vsRSV_zscore_inputCarnival <- 
  t(progeny(as.matrix(dds_A549vsRSV_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = FALSE))
colnames(pathways_A549vsRSV_zscore_inputCarnival) <- "Activity"
```

We now display the normalized enrichment scores (NES) in a bar plot. 

```{r, message=FALSE}
pathways_A549vsRSV_zscore_df <- as.data.frame(pathways_A549vsRSV_zscore) %>% 
    rownames_to_column(var = "Pathway") %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(Pathway = factor(Pathway))
```
```{r, message=FALSE, dpi=300}
ggplot(pathways_A549vsRSV_zscore_df,aes(x = reorder(Pathway, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways")
```

### A549 mock treated vs infected with HPIV3

We perform the progeny activity analysis for A549 cell line infected with HPIV3

We first compute **Progeny** scores per every sample (with the replicates) using 
the normalised counts. We display the results in a Heatmap.

```{r}
pathways_A549vsHPIV3_counts <- progeny(norm_counts_A549vsHPIV3, scale=TRUE, 
    organism="Human", top = 100)
activities_A549vsHPIV3 <- as.vector(pathways_A549vsHPIV3_counts)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

progenyBreaks <- c(seq(min(activities_A549vsHPIV3), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(activities_A549vsHPIV3)/paletteLength, 
    max(activities_A549vsHPIV3), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
progeny_hmap <- pheatmap(t(pathways_A549vsHPIV3_counts),fontsize=14, 
    fontsize_row = 10, fontsize_col = 10, 
    color=myColor, breaks = progenyBreaks, 
    main = "PROGENy (100)", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

We can see that the different replicates cluster together and clear different
activity patterns between the infected and the mock treated A549 lines. To 
evaluate the significance and sign of these pathway activities, we can run again
**Progeny** using the statistic from the differential express analysis.

```{r, message=FALSE}
dds_A549vsHPIV3_df <- as.data.frame(dds_A549vsHPIV3) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") 

pathways_A549vsHPIV3_zscore <- t(progeny(as.matrix(dds_A549vsHPIV3_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = TRUE))
colnames(pathways_A549vsHPIV3_zscore) <- "NES"

## I also need to run progeny in such a way to have values between 1 and -1 to
## use as CARNIVAL input
pathways_A549vsHPIV3_zscore_inputCarnival <- 
  t(progeny(as.matrix(dds_A549vsHPIV3_df), 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = FALSE))
colnames(pathways_A549vsHPIV3_zscore_inputCarnival) <- "Activity"
```

We now display the normalized enrichment scores (NES) in a bar plot. 

```{r, message=FALSE}
pathways_A549vsHPIV3_zscore_df <- as.data.frame(pathways_A549vsHPIV3_zscore) %>% 
    rownames_to_column(var = "Pathway") %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(Pathway = factor(Pathway))
```
```{r, message=FALSE, dpi=300}
ggplot(pathways_A549vsHPIV3_zscore_df,aes(x = reorder(Pathway, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways")
```

### Comparison: Differential infection viral effects on Pathways activity  

We finally plot together the pathway activities under the three different 
viral infections. We include in the comparison the results from the 
previous analysis (A549 mock treated vs SARS-CoV-infected Series 2)

```{r}
pathways_A549vsCOV2_zscore_df_series2 <- 
    readRDS("../IntermediateFiles/pathways_A549vsCOV2_zscore.rds") %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "Pathway") %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(Pathway = factor(Pathway))

AllInfections_Pathways <- bind_rows(pathways_A549vsCOV2_zscore_df,
    pathways_A549vsRSV_zscore_df, pathways_A549vsHPIV3_zscore_df,
    pathways_A549vsCOV2_zscore_df_series2) %>% 
    tibble::add_column(
        Virus = c(rep("SARS-CoV-2_S5",14),rep("RDS",14), 
                  rep("HPIV3",14), rep("SARS-CoV-2_S2",14)))
```
```{r, dpi=300}
ggplot(AllInfections_Pathways,aes(x = Pathway, y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    facet_wrap(~Virus, ncol = 2, nrow = 2) +
    scale_fill_gradient2(low = "darkblue", high = "darkred", 
        mid = "white", midpoint = 0) + 
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =8, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways") 
```


## Transcription Factor activity with Dorothea and Viper

Now, we estimate the transcription factor (TF) activity using the **dorothea** 
package. We select Dorothea interactions with confidence level A,B and C. 

```{r}
## We load Dorothea Regulons
data(dorothea_hs, package = "dorothea")
regulons <- dorothea_hs %>%
  dplyr::filter(confidence %in% c("A", "B","C"))
```

It is to note that for **Dorothea**, we proceed the other way around than for 
**Progeny**. We have many TFs so, we cannot visualize all of them in the same 
heatmap. That is why, we first compute the TF activity using the statistcs from 
the differential expression analysis. This will allows us to select the TFs 
whose activity varies the most between the mock treated and infected samples. 

### A549 mock treated vs infected with SARS-CoV-2 

We first run Viper using the statistic from the different expression analysis.

```{r}
dds_A549vsCOV2_stat <-  as.data.frame(dds_A549vsCOV2) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549vsCOV2_stat <- 
    dorothea::run_viper(as.matrix(dds_A549vsCOV2_stat), regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, nes = TRUE))
```

We now display the top 25 normalized enrichment scores (NES) for the TF in a 
bar plot. 

```{r}
tf_activities_A549vsCOV2_top25 <- tf_activities_A549vsCOV2_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(25, wt = abs(NES)) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))
```   
```{r, dpi=300}
ggplot(tf_activities_A549vsCOV2_top25,aes(x = reorder(GeneID, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Transcription Factors")
```

We now compute TFs activities per every sample (with the replicates) using 
the normalised counts. We display the results of the previous 25 TFs in a 
Heatmap.

```{r}
tf_activities_A549vsCOV2_counts <- 
    dorothea::run_viper(norm_counts_A549vsCOV2, regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, method = c("scale")))

tf_activities_A549vsCOV2_counts_filter <- tf_activities_A549vsCOV2_counts %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::filter(GeneID %in% tf_activities_A549vsCOV2_top25$GeneID) %>%
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549vsCOV2 <- as.vector(tf_activities_A549vsCOV2_counts_filter)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

dorotheaBreaks <- c(seq(min(tf_activities_A549vsCOV2), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(tf_activities_A549vsCOV2)/paletteLength, 
    max(tf_activities_A549vsCOV2), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
dorothea_hmap <- pheatmap(tf_activities_A549vsCOV2_counts_filter,
    fontsize=14, fontsize_row = 8, fontsize_col = 8, 
    color=myColor, breaks = dorotheaBreaks,
    main = "Dorothea ABC", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

### A549 mock treated vs infected with RSV

We first run Viper using the statistic from the different expression analysis.

```{r}
dds_A549vsRSV_stat <-  as.data.frame(dds_A549vsRSV) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549vsRSV_stat <- 
    dorothea::run_viper(as.matrix(dds_A549vsRSV_stat), regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, nes = TRUE))
```

We now display the top 25 normalized enrichment scores (NES) for the TF in a 
bar plot. 

```{r}
tf_activities_A549vsRSV_top25 <- tf_activities_A549vsRSV_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(25, wt = abs(NES)) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))
```   
```{r, dpi=300}
ggplot(tf_activities_A549vsRSV_top25,aes(x = reorder(GeneID, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Transcription Factors")
```

We now compute TFs activities per every sample (with the replicates) using 
the normalised counts. We display the results of the previous 25 TFs in a 
Heatmap.

```{r}
tf_activities_A549vsRSV_counts <- 
    dorothea::run_viper(norm_counts_A549vsRSV, regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, method = c("scale")))

tf_activities_A549vsRSV_counts_filter <- tf_activities_A549vsRSV_counts %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::filter(GeneID %in% tf_activities_A549vsRSV_top25$GeneID) %>%
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549vsRSV <- as.vector(tf_activities_A549vsRSV_counts_filter)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

dorotheaBreaks <- c(seq(min(tf_activities_A549vsRSV), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(tf_activities_A549vsRSV)/paletteLength, 
    max(tf_activities_A549vsRSV), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
dorothea_hmap <- pheatmap(tf_activities_A549vsRSV_counts_filter,
    fontsize=14, fontsize_row = 8, fontsize_col = 8, 
    color=myColor, breaks = dorotheaBreaks,
    main = "Dorothea ABC", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```


### A549 mock treated vs infected with HPIV3

We first run Viper using the statistic from the different expression analysis.

```{r}
dds_A549vsHPIV3_stat <-  as.data.frame(dds_A549vsHPIV3) %>% 
    rownames_to_column(var = "GeneID") %>% 
    dplyr::select(GeneID, stat) %>% 
    dplyr::filter(!is.na(stat)) %>% 
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549vsHPIV3_stat <- 
    dorothea::run_viper(as.matrix(dds_A549vsHPIV3_stat), regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, nes = TRUE))
```

We now display the top 25 normalized enrichment scores (NES) for the TF in a 
bar plot. 

```{r}
tf_activities_A549vsHPIV3_top25 <- tf_activities_A549vsHPIV3_stat %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(25, wt = abs(NES)) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))
```   
```{r, dpi=300}
ggplot(tf_activities_A549vsHPIV3_top25,aes(x = reorder(GeneID, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Transcription Factors")
```

We now compute TFs activities per every sample (with the replicates) using 
the normalised counts. We display the results of the previous 25 TFs in a 
Heatmap.

```{r}
tf_activities_A549vsHPIV3_counts <- 
    dorothea::run_viper(norm_counts_A549vsHPIV3, regulons,
    options =  list(minsize = 5, eset.filter = FALSE, 
    cores = 1, verbose = FALSE, method = c("scale")))

tf_activities_A549vsHPIV3_counts_filter <- tf_activities_A549vsHPIV3_counts %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::filter(GeneID %in% tf_activities_A549vsHPIV3_top25$GeneID) %>%
    column_to_rownames(var = "GeneID") %>%
    as.matrix()

tf_activities_A549vsHPIV3 <- as.vector(tf_activities_A549vsHPIV3_counts_filter)

paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

dorotheaBreaks <- c(seq(min(tf_activities_A549vsHPIV3), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(tf_activities_A549vsHPIV3)/paletteLength, 
    max(tf_activities_A549vsHPIV3), 
    length.out=floor(paletteLength/2)))
```
```{r, dpi=300}
dorothea_hmap <- pheatmap(tf_activities_A549vsHPIV3_counts_filter,
    fontsize=14, fontsize_row = 8, fontsize_col = 8, 
    color=myColor, breaks = dorotheaBreaks,
    main = "Dorothea ABC", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```

### Comparison: Differential infection viral effects on TF activity  

We finally plot together the TF activities under the three different 
viral infections. Again we also include the data from the previous A549 mock
treated vs SARS-CoV-2 comparison (series2)

```{r, message=FALSE, warning=FALSE}
tf_activities_A549vsCOV2_top25_series2 <- 
    readRDS("../IntermediateFiles/tf_activities_A549vsCOV2_stat.rds") %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "stat") %>%
    dplyr::top_n(25, wt = abs(NES)) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))

AllInfections_TFs <- bind_rows(tf_activities_A549vsCOV2_top25,
    tf_activities_A549vsRSV_top25, tf_activities_A549vsHPIV3_top25,
    tf_activities_A549vsCOV2_top25_series2) %>% 
    tibble::add_column(
        Virus = c(rep("SARS-CoV-2_S5",25),rep("RDS",25),
            rep("HPIV3",25),rep("SARS-CoV-2_S2",25)))
```
```{r, dpi=300}
ggplot(AllInfections_TFs,aes(x = GeneID, y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    facet_wrap(~Virus,drop= FALSE, nrow = 2, ncol=2) +
    scale_fill_gradient2(low = "darkblue", high = "darkred", 
        mid = "white", midpoint = 0) + 
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 90, hjust = 1, size =5),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) 
```


### Saving Results 

```{r, eval = FALSE}
saveRDS(pathways_A549vsCOV2_counts, 
    file = "IntermediateFiles/pathways_A549vsCOV2_counts.rds")
saveRDS(pathways_A549vsCOV2_zscore,
    file = "IntermediateFiles/pathways_A549vsCOV2_zscore.rds")
saveRDS(pathways_A549vsCOV2_zscore_inputCarnival,
    file = "IntermediateFiles/pathways_A549vsCOV2_zscore_inputCarnival.rds")
saveRDS(tf_activities_A549vsCOV2_stat, 
    file = "IntermediateFiles/tf_activities_A549vsCOV2_stat.rds")
saveRDS(tf_activities_A549vsCOV2_counts, 
    file = "IntermediateFiles/tf_activities_A549vsCOV2_counts.rds")

saveRDS(pathways_A549vsRSV_counts, 
    file = "IntermediateFiles/pathways_A549vsRSV_counts.rds")
saveRDS(pathways_A549vsRSV_zscore,
    file = "IntermediateFiles/pathways_A549vsRSV_zscore.rds")
saveRDS(pathways_A549vsRSV_zscore_inputCarnival,
    file = "IntermediateFiles/pathways_A549vsRSV_zscore_inputCarnival.rds")
saveRDS(tf_activities_A549vsRSV_stat, 
    file = "IntermediateFiles/tf_activities_A549vsRSV_stat.rds")
saveRDS(tf_activities_A549vsRSV_counts, 
    file = "IntermediateFiles/tf_activities_A549vsRSV_counts.rds")

saveRDS(pathways_A549vsHPIV3_counts, 
    file = "IntermediateFiles/pathways_A549vsHPIV3_counts.rds")
saveRDS(pathways_A549vsHPIV3_zscore,
    file = "IntermediateFiles/pathways_A549vsHPIV3_zscore.rds")
saveRDS(pathways_A549vsHPIV3_zscore_inputCarnival,
    file = "IntermediateFiles/pathways_A549vsHPIV3_zscore_inputCarnival.rds")
saveRDS(tf_activities_A549vsHPIV3_stat, 
    file = "IntermediateFiles/tf_activities_A549vsHPIV3_stat.rds")
saveRDS(tf_activities_A549vsHPIV3_counts, 
    file = "IntermediateFiles/tf_activities_A549vsHPIV3_counts.rds")
```


## Session Info Details

```{r, echo=FALSE, eval=TRUE}
sessionInfo()
```

